Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-22 manuscript/main.qmd
remove en-dash spaces

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples originating from near the ocean surface or mixed layer. Samples labelled as "thermocline" or "sub-thermocline" were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained. For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43&deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene -- early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

@@

<!-- REMOVED
Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`&deg;C from the equator to the poles. When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` -- `r modern_eocene_gradient$"97.5%"`)&deg;C. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`&deg;C (95 % CI).

@@

modelling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST). The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`&deg;C when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` -- `r moeotemp_global$"97.5%"`)&deg;C. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`&deg;C.
-->

@@

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig. 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)&deg;C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig. S2). In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig. S3).

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)&deg;C. The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S4).

@@

Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data with moderate sample sizes (*n* = 10 -- 34) and patchy sampling distributions (Fig. 2). This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g. @Bijl2009; @Tierney2017], which can fit only specific types of gradients. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020]. This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns, without the option of including additional knowledge (e.g. that latitudinal temperature gradients should be broadly negative).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg. Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]

@@

Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 -- 37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including TEX~86~ [@Inglis2020]. Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5&deg;C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`&deg;C.

>>>>

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer. Samples labelled as "thermocline" or "sub-thermocline" were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained. For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N--32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43&deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

@@

<!-- REMOVED
Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig. 3b). The 95% credible interval (CI) of reconstructed gradients (Fig. 3b) is noticeably narrower than the spread of empirical temperature values (Fig. 3a). The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`&deg;C from the equator to the poles. When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"`--`r modern_eocene_gradient$"97.5%"`)&deg;C. A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`&deg;C (95 % CI).

@@

modelling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST). The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`&deg;C when using the full, modern data set. When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"`--`r moeotemp_global$"97.5%"`)&deg;C. The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`&deg;C.
-->

@@

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig. 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`)&deg;C higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`)&deg;C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)&deg;C--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"`--`r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"`--`r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig. S2). In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig. S3).

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"`--`r eeco_south_north_global_diff$"97.5%"`)&deg;C. The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"`--`r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S4).

@@

Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data with moderate sample sizes (*n* = 10--34) and patchy sampling distributions (Fig. 2). This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g. @Bijl2009; @Tierney2017], which can fit only specific types of gradients. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020]. This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns, without the option of including additional knowledge (e.g. that latitudinal temperature gradients should be broadly negative).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg. Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]

@@

Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4--37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including TEX~86~ [@Inglis2020]. Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5&deg;C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`&deg;C.
